Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota

### 2010-02-28 Paper.tex
edited beginning and bayesian statistics

    \subsection{Bayesian statistics}
    Bayes' Theorem: \(P(\theta|X) = P(X|\theta)P(\theta) / P(X)\) \\

    This states that the posterior distribution \(\propto\) the likelihood
    \(\times\) the prior. Usually, the mean of the posterior distribution is
    then used as a point estimate for the parameter.

    We can use Bayesian statistics in the Hierarchical model as long as we set a
    prior for every parameter that we include. We can then use Markov Chain
    Monte Carlo (MCMC) to find a posterior distributions for every parameter. We
    run into a problem of duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is the same
    as \(p=y,E=x,\psi=1-z\). Since we know that is reasonable for \(p\) to be
    greater than \(E\), we solved this problem by restricting the distribution
    of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to \(0\text{ to }0.4\),  
    an approach recommended by Royle and Link \cite{Royle2006}.

@@

    \subsection{Simulating spatially correlated data}
    Because we modeled spatial correlation in $\theta$, we began by simulating
    spatial correlation in the track-laying process. To do this, we added an
    additional variable to the simulated data, $\alpha$. Under the new
    simulation method, $\theta$ is the probability of an otter laying down a
    track in site $i$ given site $i$ is occupied and site $i-1$ does
    \textit{not} have a track. If site $i$ is occupied and site $i-1$ does have
    a track, then the probability of an otter laying down a track in site $i$ is
    equal to $\theta+\alpha$. This means if there is a track in one site, there
    is more likely to be a track in the next site. However, when we ran data
    simulated using this new method on the model described in section \ref{car
    model}, the model almost never converged. We tested independence of the data
    using the $\hat{s}$ statistic we developed (section \ref{shat}) and found
    that the simulated data displayed much more independence than the actual
    data. We tried modeling spatial correlation in the occupancy process, $\psi
    $, instead of in the track-laying process, but the simulated data were again 
    more independent than the real data.

>>>>

    \subsection{Bayesian statistics}
	Most occupancy models utitlize maximum likelihood estimation, but we chose 
	to use a Bayesian approach to model the data for several reasons. 
	First, the occupancy models developed using maximum likelihood methods 
	assume that the sampled locations are a random sample from a large, 
	theoretically infinite, population.  In this study, the entire area of 
	interest was observed, so this assumption does not hold true.  Bayesian 
	models do not make such an assumption.  Second, because our model added extra
	parameters, formulating the likelihood function was cumbersome and maximizing
	it would have been difficult.  We added an additional random process, track 
	laying, and also included the possibility of false detection.  A Bayesian
	model seems much more intuitive and natural.

    Once we had a model formulated, we could then use Markov Chain Monte Carlo 
	(MCMC) to find a posterior distributions for every parameter. Due to the 
	nature of occupancy models, certain models can have multiple interpretations.
	We run into a problem of duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is 
	equivalent in terms of likelihood to \(p=y,E=x,\psi=1-z\) due to the 
	symmetry of the model \cite{Royle2006}.  Since we know that is reasonable 
	for \(p\) to be greater than \(E\), we solved this problem by restricting 
	the distribution of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to 
	\(0\text{ to }0.4\), an adaptation of the approach recommended by Royle and 
	Link \cite{Royle2006}.

@@

    \subsection{Simulating spatially correlated data}
    Because we modeled spatial correlation in $\theta$, we began by simulating
    spatial correlation in the track-laying process. To do this, we added an
    additional variable to the simulated data, $\alpha$. Under the new
    simulation method, $\theta$ is the probability of an otter laying down a
    track in site $i$ given site $i$ is occupied and site $i-1$ does
    \textit{not} have a track. If site $i$ is occupied and site $i-1$ does have
    a track, then the probability of an otter laying down a track in site $i$ is
    equal to $\theta+\alpha$. This means if there is a track in one site, there
    is more likely to be a track in the next site. However, when we ran data
    simulated using this new method on the model described in section \ref{car
    model}, the model almost never converged. We tested independence of the data
    using the $\hat{s}$ statistic we developed (section \ref{shat}) and found
    that the simulated data displayed much more independence than the actual
    data. We would like our simulated data to have the same level of independence 
	as our actual data.  We tried modeling spatial correlation in the occupancy 
	process, $\psi$, instead of in the track-laying process, but the simulated 
	data were again more independent than the real data.
